Ownership types in theory and practice (in Rust)

### 2023-01-25 paper/paper.tex
Worked some more on slidesmake

\begin{lstlisting}[language=C++,captionpos=b,caption={C++ use-after-free}]
std::vector<int> v { 10, 11 };
int *vptr = &v[1]; // points into v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free}]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points into v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust lifetime annotations}]
fn main() {
  let s1 = String::from("short");
  let s2 = String::from("longest");
  let longest = longest(&s1, &s2);
  println!("longest string: '{}'", longest);
}
fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
  if s1.len() > s2.len() {
    s1
  } else {
    s2
  }
}
\end{lstlisting}
If we omit the lifetime annotations \verb|'a|, the compiler will tell us that the return type of the \verb|longest| function contains a borrowed value, but it is unclear whether it is borrowed from \emph{s1} or \emph{s2}.
In other words, the compiler requires that \emph{s1} and \emph{s2} have to live at least as long as lifetime \verb|'a|.

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free with lifetimes}]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // lifetime 'a starts
Vec::push(&mut v, 12); // lifetime 'b
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for a successful compilation of the program.
\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free corrected}]
let mut v = vec![10, 11];
Vec::push(&mut v, 12); // lifetime 'b
let vptr = &mut v[1];  // lifetime 'a starts
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

>>>>

\begin{lstlisting}[language=C++,captionpos=b,caption={C++ use-after-free \cite{understanding-evolving-rust}}]
std::vector<int> v { 10, 11 };
int *vptr = &v[1]; // points into v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free \cite{understanding-evolving-rust}}]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points into v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust lifetime annotations \cite{rust-book}}]
fn main() {
  let s1 = String::from("short");
  let s2 = String::from("longest");
  let longest = longest(&s1, &s2);
  println!("longest string: '{}'", longest);
}
fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
  if s1.len() > s2.len() {
    s1
  } else {
    s2
  }
}
\end{lstlisting}
If we omit the lifetime annotations \verb|'a|, the compiler will tell us that the return type of the \verb|longest| function contains a borrowed value, but it is unclear whether it is borrowed from \emph{s1} or \emph{s2}.
In other words, the compiler requires that \emph{s1} and \emph{s2} have to live at least as long as lifetime \verb|'a|.

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free with lifetimes \cite{understanding-evolving-rust}}]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // lifetime 'a starts
Vec::push(&mut v, 12); // lifetime 'b
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for a successful compilation of the program.
\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free corrected \cite{understanding-evolving-rust}}]
let mut v = vec![10, 11];
Vec::push(&mut v, 12); // lifetime 'b
let vptr = &mut v[1];  // lifetime 'a starts
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

### 2023-01-25 slides/slides.tex
Worked some more on slidesmake

\documentclass{beamer}

\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{listings-rust}
\lstset{
  frame=tb,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle=\footnotesize\ttfamily,
  numbers=left,
  numbersep=5pt,
  numberstyle=\color{gray},
  keywordstyle=\bfseries\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=false,
  tabsize=2
}
\usepackage{tikz}
\usetikzlibrary{fit,positioning}
\definecolor{tumblue}{RGB}{0,101,189}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{cgreen}{RGB}{60,201,65}
\setbeamerfont{footnote}{size=\tiny}
\renewcommand\footnoterule{}
\usecolortheme{orchid}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{sidebar right}{% also implies no navbar
  \llap{\tikz\fill[tumblue,scale=.1] (0,0) ++(-5cm,-5cm) ++(-10cm,0)
      +(0cm,0cm) -- +(4cm,0cm) -- +(4cm,-4cm) -- +(5cm,-4cm) -- +(5cm,0cm) --
      +(10cm,0cm) -- +(10cm,-5cm) -- +(9cm,-5cm) -- +(9cm,-1cm) --
      +(8cm,-1cm) -- +(8cm,-5cm) -- +(7cm,-5cm) -- +(7cm,-1cm) --
      +(6cm,-1cm) -- +(6cm,-5cm) -- +(3cm,-5cm) -- +(3cm,-1cm) --
      +(2cm,-1cm) -- +(2cm,-5cm) -- +(1cm,-5cm) -- +(1cm,-1cm) --
      +(0cm,-1cm) -- cycle;}
  % Slide number in triangle.
  \vfill\llap{\tikz\path[fill=tumblue,inner sep=3pt] (0,0) -- ++(-1cm,0) -- +(1cm,1cm) -- cycle node[anchor=south east] {\usebeamerfont{footline}\color{white}\bfseries\insertframenumber};}%
  % Variant without triangle
  % \vfill\llap{\usebeamerfont{footline}\usebeamercolor[fg]{footline}\insertframenumber\hskip3pt}\vskip3pt%
}

@@

\begin{frame}{Memory safety}
\begin{itemize}
  \item definition: program is memory safe if all memory pointers or references always refer to valid memory
  \item Google and Microsoft: 70 percent of vulnerabilities due to memory safety issues
  \item National Security Agency (USA) recommends memory-safe languages (instead of C/C++)
  \item examples: buffer overflows, memory leaks, double-free, use-after-free
  \item result: malicious exploits, incorrect program results, "random" crashes
\end{itemize}
\end{frame}

\begin{frame}{Aliasing}
\begin{itemize}
  \item object aliasing: accessing the same memory through different symbolic names
  % TODO: example
  \item how to prevent bugs through (unintentional) aliases $\rightarrow$ bugs (hard to detect)
  \begin{itemize}
    \item ban aliasing altogether?
    \item restrict where and how aliasing can be used (through ownership rules) $\rightarrow$ improve memory safety
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Aliasing: example}
C++
\begin{lstlisting}[language=C++]
std::vector<int> v { 10, 11 };
int *vptr = &v[1];  // points into v
v.push_back(12);    // v buffer reallocated => vptr dangling
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

@@

\section{Ownership types in theory}

\begin{frame}{Ownership types}
\begin{itemize}
  \item object-oriented programs:
  \begin{itemize}
    \item objects can reference any other object
    \item objects can read and modify other objects' fields
    \item makes it difficult to understand, to maintain, and to reason about
  \end{itemize}
  \item using ownership types:
  \begin{itemize}
    \item limit which objects can be referenced
    \item can specify whether the referenced objects may be mutated or just read from
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Ownership types: core concepts}
\begin{itemize}
  % TODO: different highlighting than italics
  \item objects can be \emph{owners} of other objects
  \item \emph{inside} relation: \emph{a} is \emph{inside} \emph{b} if they are the same object or \emph{a} is transitively owned by \emph{b}
  \item \emph{outside} relation: converse of \emph{inside} relation
  \item \emph{ownership context}: set of all objects that an object owns
  \item \emph{siblings}: objects in the same ownership context
  \item special \emph{world} object is root of the ownership hierarchy tree
\end{itemize}

@@

\begin{frame}{Owners-as-dominators: implications}
\begin{itemize}
  \item any external reference to an object must go through its owner
  \item objects are only protected from external access, not internal access
  \item owners-as-dominators: owner is dominator for all of its owned objects
  \item doesn't allow common idioms that involve aliasing
  \item model restricts \emph{where} references can point, not \emph{how} references are used
\end{itemize}

@@

\begin{frame}{From theory to implementation}
\begin{itemize}
  \item Rust's ownership system: extends owners-as-modifiers
  \begin{itemize}
    \item no garbage collection (overhead)
    \item immutable (read-only) and mutable references % (not in owners-as-dominators)
    \item rules for compiler to check validity of references
    \item extensions:
    \begin{itemize}
      \item transferring ownership from one object to another
      \item object immutability by default
      \item Resource Acquisition Is Initialization (similar to C++)
    \end{itemize}
  \end{itemize}
  % \item aliasing concept essential to ownership types
  % \item different ways of implementing aliasing:
  % \begin{itemize}
  %   \item pointers: exist in Rust, but their use is not advised
  %   \item \emph{references}
  % \end{itemize}
  \item ownership rules in Rust:
  \begin{enumerate}
    \item Each value in Rust has an owner.
    \item There can only be one owner at a time.
    \item When the owner goes out of scope, the value will be \emph{dropped} (associated memory is freed).
  \end{enumerate}
\end{itemize}
\end{frame}

@@

  \begin{lstlisting}[language=Rust]
  fn main() {
    let a = String::from("hello world");
    let b = a; // ownership transfer: a is moved into b
    let c = take_ownership(b); // b is moved into function
    println!("{}", b); // compiler error: b no longer valid
  }
  fn take_ownership(s: String) -> String {
    // do something with s
    String::from("new hello world")
  }
  \end{lstlisting}
  \item create deep copies with explicit \verb|clone()| call
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Transferring ownership}
Stack-based primitives: 
\begin{itemize}
  \item small, known size at compile time
  \item examples: i32, u8, f64, char, bool etc.
  \item values are copied by default

  \begin{lstlisting}[language=Rust]
  let x: u32 = 42;
  let y: u32 = x; // x is copied instead of moved
  println!("x: {}, y: {}", x, y); // x and y valid
  \end{lstlisting}
  \item stack-based values implement \verb|Copy| trait (exclusive to \verb|Drop| trait)
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Borrowing}
\begin{itemize}
  \item \emph{reference}: an address that points to a value that is owned by another variable
  \item unlike \emph{pointers}, a reference is guaranteed to point to a valid value
  \item creating a reference in Rust is called \emph{borrowing}
  \item a reference does not own the value it points to $\rightarrow$ only reference (not the borrowed value) is dropped at the end of the reference's scope
  \item used for performing operations on values without taking ownership
  % \item the scope of a reference starts where it is introduced and extends until the last time it is used
  \item Rust supports immutable (\verb|&|) and mutable (\verb|&mut|) references
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Lifetimes}
\begin{itemize}
  \item every reference has a \emph{lifetime}
  \item references grant temporary access to a data structure $\rightarrow$ how long is "temporary"?
  \item construct that the \emph{borrow checker} uses to ensure that all borrows are valid
  \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
  \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
  \item compiler constraints:
  \begin{enumerate}
    \item The reference can only be used while its lifetime is ongoing, and
    \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
  \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Borrowing: immutable references}
\begin{itemize}
  \item read-only/shared references $\rightarrow$ compiler disallows modifying borrowed value
  \item \verb|&| and \verb|*| operators
  \item unlimited amount of immutable references to same value are allowed (if no mutable references)

@@

\begin{frame}[fragile]{Borrowing: mutable references}
\begin{itemize}
  \item \emph{data races}: multiple pointers access the same data simultaneously and at least one is writing to the data
  \item goal: prevent \emph{data races} at compile time
    % TODO: latex beamer make stand out/highlight
  \item lifetime of mutable reference may not overlap with lifetime of any other (immutable or mutable) reference
  \item only one mutable reference to a value allowed at any given time

@@

  \item Will it compile? \pause No.
  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'.
  \item Refactored version that compiles:\pause

@@

\begin{frame}[fragile]{Borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let r1 = &s;
  let r2 = &s;
  println!("{} and {}", r1, r2);
  let r3 = &mut s;
  r3.push_str(" world");
  \end{lstlisting}

  \item Will it compile? \pause Yes.
  \item Why? \pause Lifetimes of immutable and mutable references don't overlap.

  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let r1 = &s; // lifetime 'a starts
  let r2 = &s; // lifetime 'b starts
  println!("{} and {}", r1, r2);
  // lifetimes 'a and 'b end
  let r3 = &mut s; // lifetime 'c starts
  r3.push_str(" world");
  \end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let r = &s;
  s.push_str(" world");
  println!("{}", *r);
  \end{lstlisting}

  \item Will it compile? \pause No.
  \item Why not? \pause Lifetimes of mutable and immutable references overlap (cannot borrow 's' as mutable because it is also borrowed as immutable)
  \item Refactored version that compiles:\pause

  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  s.push_str(" world"); // lifetime 'a starts and ends
  let r = &s; // lifetime 'b starts
  println!("{}", *r); // lifetime 'b ends
  \end{lstlisting}
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Another implementation: C++}
\begin{itemize}
  \item integration of move semantics into language: Rust early on, C++ late in its life
  \item move semantics closely related to constructors and assignment operators (overloaded for classes)
  \item different reference types based on value categories:
  \begin{itemize}
    \item L-value references: refer to named values $\rightarrow$ copy
    \item R-value references: refer to temporary results of expression evaluations $\rightarrow$ move
  \end{itemize}
  \item \emph{overload resolution process} will select matching overload based on reference's value category
  \begin{lstlisting}[language=C++]
  struct S {
    S() {} // default constructor
    S(const S&) {...} // copy constructor
    S(S&&) {...} // move constructor
    S operator=(const S&) {...} // copy assignment operator
    S operator=(S&&) {... } // move assignment operator
    // ... more functions and data fields etc.
  }
  \end{lstlisting}
\end{itemize}
\end{frame}

@@

\section{Extra slides}

\begin{frame}{Lifetime elision}
Here are the three rules:
\begin{itemize}
  \item Each elided lifetime in a function’s arguments becomes a distinct lifetime parameter.
  \item If there is exactly one input lifetime, elided or not, that lifetime is assigned to all elided lifetimes in the return values of that function.
  \item If there are multiple input lifetimes, but one of them is &self or &mut self, the lifetime of self is assigned to all elided output lifetimes.
\end{itemize}
Otherwise, it is an error to elide an output lifetime.
\end{frame}

>>>>

% \documentclass[handout]{beamer}
\documentclass{beamer}

\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{listings-rust}
\lstset{
  frame=tb,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle=\footnotesize\ttfamily,
  numbers=left,
  numbersep=5pt,
  numberstyle=\color{gray},
  keywordstyle=\bfseries\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=false,
  tabsize=2
}
\usepackage{tikz}
\usetikzlibrary{fit,positioning}
\definecolor{tumblue}{RGB}{0,101,189}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{cgreen}{RGB}{60,201,65}
\setbeamerfont{footnote}{size=\tiny}
\renewcommand\footnoterule{}
\usecolortheme{orchid}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{sidebar right}{% also implies no navbar
  \llap{\tikz\fill[tumblue,scale=.1] (0,0) ++(-5cm,-5cm) ++(-10cm,0)
      +(0cm,0cm) -- +(4cm,0cm) -- +(4cm,-4cm) -- +(5cm,-4cm) -- +(5cm,0cm) --
      +(10cm,0cm) -- +(10cm,-5cm) -- +(9cm,-5cm) -- +(9cm,-1cm) --
      +(8cm,-1cm) -- +(8cm,-5cm) -- +(7cm,-5cm) -- +(7cm,-1cm) --
      +(6cm,-1cm) -- +(6cm,-5cm) -- +(3cm,-5cm) -- +(3cm,-1cm) --
      +(2cm,-1cm) -- +(2cm,-5cm) -- +(1cm,-5cm) -- +(1cm,-1cm) --
      +(0cm,-1cm) -- cycle;}
  % Slide number in triangle.
  \vfill\llap{\tikz\path[fill=tumblue,inner sep=3pt] (0,0) -- ++(-1cm,0) -- +(1cm,1cm) -- cycle node[anchor=south east] {\usebeamerfont{footline}\color{white}\bfseries\insertframenumber};}%
  % Variant without triangle
  % \vfill\llap{\usebeamerfont{footline}\usebeamercolor[fg]{footline}\insertframenumber\hskip3pt}\vskip3pt%
}
% TODO: navigation bar
% \setbeamertemplate{footline}{%
%  \begin{beamercolorbox}[ignorebg,ht=2.25ex,dp=3.75ex]%
%        {section in head/foot}%
%    \insertsectionnavigationhorizontal{\dimexpr\paperwidth-1cm}{}{}%
%  \end{beamercolorbox}%
% }

@@

\begin{frame}{Memory safety}
\begin{itemize}
  \item definition: program is memory safe if all memory pointers or references always refer to valid memory
  % \item Google and Microsoft: 70 percent of vulnerabilities due to memory safety issues
  % \item National Security Agency (USA) recommends memory-safe languages (instead of C/C++)
  \item examples: buffer overflows, memory leaks, double-free, use-after-free
  \item result: malicious exploits, incorrect program results, "random" crashes
\end{itemize}
\end{frame}

% \begin{frame}{Aliasing}
% \begin{itemize}
%   \item 
%   % TODO: example
%   \item how to prevent bugs through unintentional aliases (hard to detect)?
%   \begin{itemize}
%     \item ban aliasing altogether
%     \item restrict where and how aliasing can be used (through ownership rules) $\rightarrow$ improve memory safety
%   \end{itemize}
% \end{itemize}
% \end{frame}

\begin{frame}[fragile]{Aliasing}
\emph{aliasing}: accessing same memory through different symbolic names
\pause

C++
\begin{lstlisting}[language=C++]
std::vector<int> v { 10, 11 };
int *vptr = &v[1];  // points into v
v.push_back(12);    // v buffer reallocated => vptr dangling
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}
\pause

@@

\section{Ownership types}

\begin{frame}{Ownership types}
\begin{itemize}
  % \item object-oriented programs:
  % \begin{itemize}
  %   \item objects can reference any other object
  %   \item objects can read and modify other objects' fields
  %   \item makes it difficult to understand, to maintain, and to reason about
  % \end{itemize}
  \item using ownership types:
  \begin{itemize}
    \item limit which objects can be referenced
    \item specify whether the referenced objects may be mutated or just read from
    \item no garbage collection (overhead)
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Ownership types: core concepts}
\begin{itemize}
  % TODO: different highlighting than italics
  \item objects can be \emph{owners} of other objects
  \item special \emph{world} object is root of the ownership hierarchy tree
  \item \emph{ownership context}: set of all objects that an object owns
  \item \emph{siblings}: objects in the same ownership context
  \item \emph{inside} relation: \emph{a} is \emph{inside} \emph{b} if they are the same object or \emph{a} is transitively owned by \emph{b}
  \item \emph{outside} relation: converse of \emph{inside} relation
\end{itemize}

@@

\begin{frame}{Owners-as-dominators: implications}
\begin{itemize}
  \item any external reference to an object's internals must go through its owner
  \item objects are only protected from external access, not internal access
  % \item owners-as-dominators: owner is dominator for all of its owned objects
  \item model restricts \emph{where} references can point, not \emph{how} references are used
  \item doesn't allow common idioms that involve aliasing
\end{itemize}

@@

\begin{frame}[fragile]{From theory to implementation}
\begin{itemize}
  \item Rust's ownership system: extends owners-as-modifiers
  \begin{itemize}
    \item immutable (read-only) and mutable references % (not in owners-as-dominators)
    \item rules for compiler to check validity of references
  \end{itemize}

  \pause
  \item ownership rules in Rust:
  \begin{enumerate}
    \item Each value has an owner.
    \item There can only be one owner at a time.
    \item When the owner goes out of scope, the value will be \emph{dropped} (associated memory is freed).
  \end{enumerate}
  \begin{lstlisting}[language=Rust]
  { // a is not declared yet
    let a = 42; // scope of a starts here
    // do stuff with a
  } // scope is over => a is no longer valid
  \end{lstlisting}

  \pause
  \item Variables are immutable by default
  \begin{lstlisting}[language=Rust]
  let a = 42; // immutable
  a = 24; // invalid
  let mut b = 42; // mutable
  b = 24; // valid
  \end{lstlisting}
\end{itemize}
\end{frame}

@@

  \begin{lstlisting}[language=Rust]
  fn main() {
    let a = String::from("hello world");
    let b = a; // ownership transfer: a is moved into b
    println!("{}", a); // compiler error: a no longer valid
    let c = take_ownership(b); // b is moved into function
  }
  fn take_ownership(s: String) -> String {
    // do something with s
    String::from("new hello world")
  }
  \end{lstlisting}
  \item create deep copies with explicit \verb|clone()| call
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Transferring ownership}
Stack-based primitives: 
\begin{itemize}
  \item examples: i32, u8, f64, char, bool etc.
  \item small, known size at compile time
  \item values are copied by default

  \begin{lstlisting}[language=Rust]
  let x: u32 = 42;
  let y: u32 = x; // x is copied instead of moved
  println!("x: {}, y: {}", x, y); // x and y valid
  \end{lstlisting}
  % \item stack-based values implement \verb|Copy| trait
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Borrowing}
\begin{itemize}
  \item creating a reference in Rust is called \emph{borrowing}
  \item \emph{reference}: an address that points to a value that is owned by another variable
  \item unlike \emph{pointers}, a reference is guaranteed to point to a valid value
  % \item a reference does not own the value it points to $\rightarrow$ only reference (not the borrowed value) is dropped at the end of the reference's scope
  \item used for performing operations on values without taking ownership
  \item every reference has a \emph{lifetime}: scope for which reference is valid
  \item lifetime of borrowed value must be at least as long as lifetime of reference
  % \item a reference does not own the value it points to
  % \item the scope of a reference starts where it is introduced and extends until the last time it is used
  % \item Rust supports immutable (\verb|&|) and mutable (\verb|&mut|) references
\end{itemize}
\end{frame}

% \begin{frame}[fragile]{Lifetimes}
% \begin{itemize}
%   \item references grant temporary access to values $\rightarrow$ how long is "temporary"?
%   \item every reference has a \emph{lifetime}: scope for which reference is valid
%   \item \emph{borrow checker} uses lifetimes to ensure that all borrows are valid

%   % \item program is valid if all of the below hold for all references:
%   % \begin{enumerate}
%   %   \item the reference can only be used while its lifetime is ongoing
%   %   % \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
%   %   \item immutable reference: the borrowed value does not get mutated during lifetime
%   %   \item mutable reference: the borrowed value is not used at all during lifetime
%   % \end{enumerate}

%   % \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
%   % \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
% \end{itemize}
% \end{frame}

\begin{frame}[fragile]{Borrowing: immutable references}
\begin{itemize}
  \item read-only/shared references $\rightarrow$ compiler disallows modifying borrowed value
  \item \verb|&| and \verb|*| operators
  \item borrow-checker rule: the borrowed value does not get mutated during lifetime
  \item unlimited amount of immutable references to same value (if no mutable references or mutations)

@@

\begin{frame}[fragile]{Borrowing: mutable references}
\begin{itemize}
  % \item \emph{data races}: multiple pointers access the same data simultaneously and at least one is writing to the data
  % \item goal: prevent \emph{data races} at compile time
  \item borrow-checker rule: the borrowed value is not used at all during lifetime
  % \item lifetime of mutable reference may not overlap with lifetime of any other (immutable or mutable) reference
  % \item only one mutable reference to a value allowed at any given time

@@

  \item Will it compile? \pause No.
  \item Why not? Borrowed value 'b' does not live as long as the reference 'a'.
  \item Refactored version that compiles:\pause

@@

\begin{frame}[fragile]{Borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello world");
  let r = &s;
  s = String::from("hello world again");
  println!("{}", *r);
  \end{lstlisting}

  \item Will it compile? \pause No.
  \item Why not? Violated rule: the borrowed value does not get mutated during lifetime
  \item Refactored version that compiles:\pause

  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello world");
  s = String::from("hello world again");
  let r = &s;
  println!("{}", *r);
  \end{lstlisting}
\end{itemize}
\end{frame}

% \begin{frame}[fragile]{Borrowing: examples}
% \begin{itemize}
%   \begin{lstlisting}[language=Rust]
%   let mut s = String::from("hello");
%   let r1 = &s;
%   let r2 = &s;
%   println!("{} and {}", *r1, *r2);
%   let r3 = &mut s;
%   r3.push_str(" world");
%   \end{lstlisting}

%   \item Will it compile? \pause Yes.
%   \item Why? \pause Lifetimes of immutable and mutable references don't overlap.

%   \begin{lstlisting}[language=Rust]
%   let mut s = String::from("hello");
%   let r1 = &s; // lifetime r1 starts
%   let r2 = &s; // lifetime r2 starts
%   println!("{} and {}", *r1, *r2);
%   // lifetimes r1 and r2 end
%   let r3 = &mut s; // lifetime r3 starts
%   r3.push_str(" world");
%   \end{lstlisting}
% \end{itemize}
% \end{frame}

% \begin{frame}[fragile]{Borrowing: examples}
% \begin{itemize}
%   \begin{lstlisting}[language=Rust]
%   let mut s = String::from("hello");
%   let r = &s;
%   s.push_str(" world");
%   println!("{}", *r);
%   \end{lstlisting}

%   \item Will it compile? \pause No.
%   \item Why not? \pause Lifetimes of mutable and immutable references overlap (cannot borrow 's' as mutable because it is also borrowed as immutable)
%   \item Refactored version that compiles:\pause

%   \begin{lstlisting}[language=Rust]
%   let mut s = String::from("hello");
%   s.push_str(" world"); // lifetime starts and ends
%   let r = &s; // lifetime r starts
%   println!("{}", *r); // lifetime r ends
%   \end{lstlisting}
% \end{itemize}
% \end{frame}

@@

\begin{frame}[fragile]{Another implementation: C++}
\begin{itemize}
  % \item integration of move semantics into language: Rust early on, C++ late in its life
  \item move semantics closely related to constructors and assignment operators (overloaded for classes)
  \item different reference types based on value categories:
  \begin{itemize}
    \item L-value references: refer to named values $\rightarrow$ copy
    \item R-value references: refer to temporary results of expression evaluations $\rightarrow$ move
  \end{itemize}
  \item \emph{overload resolution process} will select matching overload based on reference's value category
  \begin{lstlisting}[language=C++]
  struct S {
    S() {} // default constructor
    S(const S&) {...} // copy constructor
    S(S&&) {...} // move constructor
    S operator=(const S&) {...} // copy assignment operator
    S operator=(S&&) {... } // move assignment operator
    // ... more functions and data fields etc.
  }
  \end{lstlisting}
\end{itemize}
\end{frame}

@@

% \section{Extra slides}

\begin{frame}[fragile]{Lifetime elision}
\begin{itemize}
  \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
  \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
  \item Here are the three rules:
  \begin{itemize}
    \item Each elided lifetime in a function’s arguments becomes a distinct lifetime parameter.
    \item If there is exactly one input lifetime, elided or not, that lifetime is assigned to all elided lifetimes in the return values of that function.
    \item If there are multiple input lifetimes, but one of them is &self or &mut self, the lifetime of self is assigned to all elided output lifetimes.
  \end{itemize}
  \item Otherwise, it is an error to elide an output lifetime.
\end{itemize}
\end{frame}
